### 2016-09-02 doc/publications/plos16/plos16.tex
methods + result

In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
can have on the prediction accuracy of a model.

\section{Methods}

% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
\section{Overview of Random Forests}

@@

\subsection{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
similar to DataFrames in R.

@@

\section{Results and Discussion}

\marginpar{These are regression examples, we need some classification examples}

To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
the training data set from $n=100 \times p=1000$ to $p=1000000$.
%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
%TODO also how is the added noise playing into this ?
The random forest model recovers variables 1 and 4 as the two
most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
% can recover the variables defining a functional relationship in the presence of a substantial number of noise
% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
% to be altered from their default value when there are a large number of predictors.

% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
% the significant varaiables if the functional relationship is simple.

% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?

\begin{figure}[tbhp]
\begin{tabular}{ll}
a)& b)\\
 \includegraphics[totalheight=6cm]{./figs/tree_function.png}&
  \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
 \end{tabular}
 \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
   variable with added Gaussian noise (in red).}
  \label{figure:synth}
\end{figure}

\subsection{CursedForest processes genome-wide variant data}
%\label{comp}
%Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and
%time/mem)

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Genome \%}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
& RF - Spark ML &  &  &  \\
& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
 \hline
\multirow{3}{*}{phase3\_chr22: 2,504 x 1,103,548} &  &  &  &  \\
& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
\hline
\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
& LR - Spark ML&  & 2hr 6min & 8GB\\ 
& ranger  R   &  0.06   &     28 mins  &            num.threads=10\footnote{don't know how to get the memory used, This
  was done on dpas-03} \\ 
\hline
\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
& RF - Spark ML & &  &\\
& LR - Spark ML & & 14hr 48min & 8GB \\ 
& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. Size of the data is listed as individuals x variants.

\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

%\begin{table}
%\begin{minipage}{\textwidth}
%\begin{center}
%\begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
%\hline
%\bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
%\hline
%RF (R)\footnote{10 nodes, 500 trees, dpas03} 
 %                                  &  0.456 & 2.89 & 18GB \\ \hline
%CursedForest\footnote{something}
 %                                  & X & X & X \\ \hline
%\end{tabular}
%\medskip
%\begin{flushleft} 
%\caption{Comparion between Random Forests on R and CursedForests on a smaller
%dataset. Chromosome 22 - phase1 (2.1GB) }
%\label{table:smaller_data}
%\end{flushleft}
%\end{center}
%\end{sidewaystable}
%\end{minipage}
%\end{table}

% \begin
% \begin{table}[!ht]
% \caption{
% {\bf Performance comparison between the different Machine Learning algorithms.}}
% \begin{tabular}{|p{4.2cm}|l|l|l|l|l|l|l|}
% \hline
% \bf{Genome }  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
% \hline
%  chr 	 - phase1 (2.1GB)  &  RF (R) &  0.456 & 2.89 & 18GB \\ \hline
% \end{tabular}
% \begin{flushleft} 
% \end{flushleft}
% \label{table1}
% %\end{adjustwidth}
% \end{table}

@@

We note that it is possible for a RF model to have a very poor fit, and be useless for prediction, and still have a 
useful ranking of variable importance.

@@

%\subsection{James' dataset}
%Comparing CursedForest with feature selection + linear model (performance and time/mem)

\subsection{Conclusion}

@@

%\section*{References}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here.
%

@@

  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691

@article{Chen2012323,
title = "Random forests for genomic data analysis ",
journal = "Genomics ",
volume = "99",
number = "6",
pages = "323 - 329",
year = "2012",
note = "",
issn = "0888-7543",
doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
author = "Xi Chen and Hemant Ishwaran",
keywords = "Random forests",
keywords = "Random survival forests",
keywords = "Classification",
keywords = "Prediction",
keywords = "Variable selection",
keywords = "Genomic data analysis ",
abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features. This makes \{RF\} particularly appealing for high-dimensional genomic data analysis. In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning. "
}

>>>>

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
can have on the prediction accuracy of a model.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Method
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DENIS Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature
% >>CONDENSE
\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
\subsection{Overview of Random Forests}

@@

% <<CONDENSE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the cursed forest implementation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cursed Forest.}
% DENIS: Aidan and Piotr can you please explain here how Cursed forest is implemented
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a ``DataFrame'',
similar to DataFrames in R.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
\subsubsection{1000 Genomes}
%DENIS: Rob how was the synthetic data generated
\subsubsection{Synthetic data}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Results
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature selection 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest grows denser forests to stabilise feature importance ranking}
In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in parallel results in a stable feature importance ranking.
This concept was originally proposed by Liaw and Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to run, the ranking of the variables will stabilise with growing \texttt{ntree}.

To investigate this hypothesis for data sets with large number of feature variables we generated a synthetic dataset ...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DENIS: Rob I don't think this is relevant anymore?
% >>CONDENSE
%\marginpar{These are regression examples, we need some classification examples}
%To test this empirically we adapt the simulations developed by Genuer {\it et al.}~\cite{Genuer.et.al.2010} to include a
%much larger feature vector $p$ per sample.  The function, shown in Figure \ref{figure:synth}a, is defined on the first 5
%predictor variables. We have added Gaussian noise to the function (Figure \ref{figure:synth}b) and extended the size of
%the training data set from $n=100 \times p=1000$ to $p=1000000$.
%%TODO recovers 2 out of 5 ? I think you need to spell out why this is a good demonstrator that RF can deal with such a large number of features...
%%TODO also how is the added noise playing into this ?
%The random forest model recovers variables 1 and 4 as the two
%most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

%% Important ?  In a number of simulations (reproduced from \cite{Genuer.et.al.2010}) we see evidence that random forests
%% can recover the variables defining a functional relationship in the presence of a substantial number of noise
%% variables. See the supplementary information for full details. It is apparently that some of the model parameters,
%% such as \texttt{mtry} -- the number of randomly selected variables considered as potential splits at each node, have
%% to be altered from their default value when there are a large number of predictors.

%% It is likely that in the genomic setting, variables will have a monotonic influence on the outcome.  We have taken
%% \cite{Genuer.et.al.2010}'s tree function example and extended it to a much larger number of noise variables.  The
%% function, shown in figure \ref{figure:tree_function.png}, is defined on the first 5 predictor variables. We have added
%% Gaussian noise to the function (figure \ref{figure:noisy_tree_function.png}) and extended the size of the training
%% data set from $n=100 \times p=1000$ to $n=100 \times p=1000000$.  The random forest model recovers variables 1 and 4
%% as the two most important variables (out of 1000000). It fails to uncover the other 3 variables, $\{2,3,5\}$.

%% It would appear that even in setting with an extremely large number of noise variables, it may be possible to recover
%% the significant varaiables if the functional relationship is simple.

%% TODO can you make the caption more descriptive, e.g. what are the splits in the three and what are the numbers;
%% similarly for b) there are 100 samples (can you name the index axis samples) and why are there 9 targets (y) ?

%\begin{figure}[tbhp]
%\begin{tabular}{ll}
%a)& b)\\
%\includegraphics[totalheight=6cm]{./figs/tree_function.png}&
% \includegraphics[totalheight=6cm]{./figs/noisy_tree_function.png}
% \end{tabular}
% \captionof{figure}{{\bf Synthetic data demonstrating RF's capability to cope with $p \gg n$.} {\bf a)} The tree
%   function, where each split means XXXX. {\bf b)} A plot of the ordered $y$ variable for the tree example, and the $y$
%   variable with added Gaussian noise (in red).}
%  \label{figure:synth}
%\end{figure}
% <<CONDENSE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Feature selection application example 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Feature selection application example}
%Maybe we should have a feature selection on real data section. We could be using one of the smaller datasets used in the other papers?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest outperforms existing methods for classification problems}
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the 1000 genomes dataset.

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
\hline
%\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  &  &  \\
&&& RF - Spark ML &  &  &  \\
&&& RF - R &  0.05  & 24 min  & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes}\\
&&& ranger -R &         0.05 &        1 min 10 seconds  &          16GB  (one processor) \\
 \hline
phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
&&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
&&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
\hline
phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
&&& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
&&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
&&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
  was done on dpas-03} \\ 
\hline
phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
&&& RF - Spark ML & &  &\\
&&& LR - Spark ML & & 14hr 48min & 8GB \\ 
&&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
  ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
\hline
phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
&&& RF - Spark ML & - & - & - \\
&&& LR - Spark ML & - & - & - \\ 
%&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. 
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Sensitivity 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sensitivity of parameter choice}

@@

\section{Conclusion}

@@

@@

%  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
%    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
%  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
%    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691

%@article{Chen2012323,
%title = "Random forests for genomic data analysis ",
%journal = "Genomics ",
%volume = "99",
%number = "6",
%pages = "323 - 329",
%year = "2012",
%note = "",
%issn = "0888-7543",
%doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
%url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
%author = "Xi Chen and Hemant Ishwaran",
%keywords = "Random forests",
%keywords = "Random survival forests",
%keywords = "Classification",
%keywords = "Prediction",
%keywords = "Variable selection",
%keywords = "Genomic data analysis ",
%abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features. This makes \{RF\} particularly appealing for high-dimensional genomic data analysis. In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning. "
%}
